---
title: "Solutions to Exercises of Introduction to Statistical Learning, Chapter 9"
author: "Guillermo Martinez Dibene"
date: "16th of May, 2021"
output: html_document
---

This chapter was all about clustering and they focus on three types of clusters: principal component analysis, K-means clustering and hierarchichal clustering.

```{r Loading libraries, include = FALSE}
library(ISLR)
library(tidyverse)
```


# Exercise 9

Here we wil analyse the `USArrests` data set in which the number of `Murders`, `Assaults` and `Rape` per 100,000 people are given for the year of 1973. The percent of urban population rounded to the nearest integer is also given in the `UrbanPop` column.

```{r}
arrests <- rownames_to_column(USArrests, var = "State") %>% as_tibble()
```

### Hierarchical clustering without scaling

First we will do a hierarchical clustering of the data using complete linkage and Euclidean distance _without scaling_.

```{r out.width = '90%'}
hier_clust <- hclust(dist(select(arrests, -State)))
plot(hier_clust, labels = arrests$State, cex = 0.75, xlab = "", sub = "")
abline(h = 120, lty = "dashed")
```

We created three clusters. We now print them.

```{r}
arrests <- arrests %>% mutate(
	cluster = cutree(hier_clust, 3)
)

```

The first cluster is:
```{r}
arrests %>%
	filter(cluster == 1) %>%
	select(State)
```

The second cluster is:
```{r}
arrests %>%
	filter(cluster == 2) %>%
	select(State)
```


The third cluster:
```{r}
arrests %>%
	filter(cluster == 3) %>%
	select(State)
```

I am no expert in the location of the USA states, but it seems fairly obvious that geographical location played a big role in the clustering, despite said feature is not part of the feature matrix.

### Hierarchical clustering with scaling

We redo the previous code but scaling the variables in advance.

```{r}
arrests_std <- arrests %>%
	select(
		State:Rape #We do not need the cluster column
	) %>%
	mutate(
		across(
			.cols = -State, #Scale everything except State
			.fns = scale
		)
	)

hier_clust_std <- hclust(dist(select(arrests_std, -State)))
```

```{r out.width = '90%'}
plot(hier_clust_std, labels = arrests$State, cex = 0.75, xlab = "", sub = "")
```

Now the states seem more mixed geographically. It makes sense that different areas of the US should share different levels of crime. For instance, before scaling, there was clearly a southern-states cluster and a northern-states one. This could be due to cultural differences. Overall, it seems that not scaling the variables for hierarchical cluster produces results that are more culturally sensitive.

# Exercise 11

Here we will work with a special data set that consists of 40 tissue samples with measurements of 1,000 genes. The first 20 samples are from healthy patients and the other 20 are from a diseased group.

```{r}
genes <- read_csv(r"(Ch10Ex11.csv)", col_names = paste(rep("Patient", 40), 1:40))
hier_genes <- hclust(as.dist(1 - cor(genes)), method = "complete")
plot(hier_genes, xlab = "", sub = "Healthy: 1-20. Diseased: 21-40", ylab = "", cex = 0.75)
```

We see that the diseased and the healthy patients have very different genes.